Local simple connectedness of boundaries of hyperbolic groups

### 2020-04-01 paper.tex
Minor changes to 4.2
Wrote a conclusion section and started work on the introduction

\todo{Write introduction}

@@

Let $\delta \geq 0$ be such that the Cayley graph of $G$ with respect to $S$ is
$\delta$-hyperbolic, meaning that all geodesic triangles in the Cayley graph of
$G$ are $\delta$-thin. To ameliorate the growth of constants later on we in
fact assume that all \emph{ideal} triangles (with vertices in $G \union
\boundary G$) are $\delta$-hyperbolic; this can be guaranteed by multiplying
$\delta$ by $4$.

@@

  Let $h$ be a contracting homotopy of $\gamma_{i_1}$ in $\boundary G -
  \set\xi$ and let $C_1$ be a compact subset of $\boundary G - \set\xi$
  containing $C_0$ and the image of $h$. Let $i_2$ be so that $C_1 \subset
  U_{i_2}$.

@@

\section{Local simple connectedness}

@@

\begin{restatable}{theorem}{ddagimpliesLCd}\label{thm:ddagimpliesLCd}
  If $G$ satisfies $\ddag'_M$ and $\S_M$ for all $M$ then $\boundary G$ is
  locally simply connected.
\end{restatable}

@@

We begin the proof of \cref{thm:ddagimpliesLCd} by proving the following 
proposition, which extends the maps constructed in
Proposition~\ref{prop:i_in_dimension_1}.

@@

\Cref{thm:ddagimpliesLCd} follows from this proposition:

\ddagimpliesLCd*

@@

>>>>

\begin{enumerate}
  \item Introduce boundaries. What are they, why are they important.
  \item Kapovich's question. Not much flexibility, not very many restrictions
    either.
  \item Partial answers.
\end{enumerate}

The goal of this paper is to prove the following theorem, which restricts the
local complexity of the boundary of a hyperbolic group in terms of its global
complexity.

\begin{theorem}\label{thm:main_theorem}
  Let $G$ be a one-ended hyperbolic group. Then $\boundary G$ is locally simply
  connected if and only if $\boundary G - \set\xi$ is simply connected for any
  point $\xi \in \boundary G$.
\end{theorem}

In Section~\ref{sec:spheres} we use the Rips complex of a group $G$ to build a
sequence of simplicial complexes $(K_n)_{n \geq 0}$ that give successively
better approximations to the boundary $\boundary G$. These simplicial complexes
are spheres in the Rips complex. There exist natural simplicial maps
$p^n_m\from K_n\to K_m$ for $n - m$ large enough, and there exist natural
continuous maps $p^\infty_n\from\boundary G \to K_n$ for all $n$.  The
remainder of Section~\ref{sec:spheres} will be spent proving quantitative
statements about these maps.

In Section~\ref{sec:horoball} we prove a compactness result: given a sequence
of vertices $v_n \in K_n$, there exist group elements $g_n$ and a subcomplex
$H$ of the Rips complex such that translation by $g_n$ maps a large
neighbourhood in $K_n$ of $v_n$ onto a large neighbourhood in $H$ of $g_n\cdot
v_n$.  We may further assume that $(g_n)$ converges to a point $\xi \in
\boundary G$, and then we think of $H$ as a horoball in $G$ asymptotic to
$\xi$. We then spend the remainder of Section~\ref{sec:horoball} describing
projection maps $\boundary G - \set\xi \to H$.

In Section~\ref{sec:double_dagger} we begin by recalling the so-called
double-dagger condition $\ddag$ of Bestvina and Mess. This condition is
guaranteed to hold if $\boundary G$ does not contain a cut point, and implies
that $\boundary G$ is locally path connected. We then modify the condition to
allow for easier generalisation, and introduce a new condition $\S$, which we
will see to be related the local simple connectedness of $\boundary G$. Using
the compactness results from Section~\ref{sec:horoball} we show that this
condition is satisfied whenever we have that $\boundary G - \set\xi$ is simply
connected for any $\xi$ in $\boundary G$.

Then, in Section~\ref{sec:local_simple_connectedness} we show that our new
condition $\S$ implies the local simple connectedness of $\boundary G$. We
begin by using $\S$ to construct a sequence of maps $(i^n_\infty)_n$ from the
2-skeleton of $K_n$ to $\boundary G$. These maps can be thought of as
approximate sections of $p^\infty_n$, in that $i^n_\infty\composed p^\infty_n$
converges uniformly to the identity map $\boundary G \to \boundary G$ as $n$
tends to $\infty$. By controlling the diameter of the image under $i^n_\infty$
of a simplex in $K_n$ we are able to prove the local simple connectedness of
$\boundary G$.

Finally, in Section~\ref{sec:conclusion} we put the results of the previous
sections together to prove Theorem~\ref{thm:main_theorem}, and conclude the
paper with some questions about strengthenings and generalisations.

@@

Let $\delta \geq 0$ be such that the Cayley graph of $G$ with respect to $S$ is
$\delta$-hyperbolic, meaning that all geodesic triangles in the Cayley graph of
$G$ are $\delta$-thin. To reduce the growth of constants later on we in
fact assume that all \emph{ideal} triangles (with vertices in $G \union
\boundary G$) are $\delta$-hyperbolic; this can be guaranteed by multiplying
$\delta$ by $4$.

@@

  Let $h$ be a contracting homotopy of $\gamma_{i_1}$ in $\boundary G -
  \set\xi$ and let $C_1$ be a compact subset of $\boundary G - \set\xi$
  containing $C_0$ and the image of $h$. Let $i_2 \geq i_1$ be so that $C_1
  \subset U_{i_2}$.

@@

\section{Local simple connectedness}\label{sec:local_simple_connectedness}

@@

\begin{restatable}{theorem}{ddagimpliesLCtwo}\label{thm:ddagimpliesLCtwo}
  If $G$ satisfies $\ddag'_M$ and $\S_M$ for all $M$ then $\boundary G$ is
  locally simply connected.
\end{restatable}

@@

We begin the proof of \cref{thm:ddagimpliesLCtwo} by proving the following 
proposition, which extends the maps constructed in
Proposition~\ref{prop:i_in_dimension_1}.

@@

\Cref{thm:ddagimpliesLCtwo} follows from this proposition:

\ddagimpliesLCtwo*

@@

\section{Conclusion}\label{sec:conclusion}

The results of Sections~\ref{sec:double_dagger}
and~\ref{sec:local_simple_connectedness} together imply one direction of
Theorem~\ref{thm:main_theorem}; we now complete the proof of this theorem.

\begin{proof}[Proof of Theorem~\ref{thm:main_theorem}]
  If $G$ is a one-ended hyperbolic group such that $\boundary G - \set\xi$ is
  simply connected for any $\xi \in \boundary G$, then the condition $\S_M$ is
  satisfied for any $M$ by Proposition~\ref{prop:sufficient_condition}. It
  follows that $\boundary G$ is locally simply connected by
  Theorem~\ref{thm:ddagimpliesLCtwo}.

  For the converse, suppose that $\boundary G$ is connected and $\boundary$ 
  is locally simply connected. Let $\xi$ be a point in $\boundary G$ and let
  $\gamma\from S^1\to\boundary G - \set\xi$ be a continuous map.

  The point $\xi$ is a conical limit point for the action of $G$ on $\boundary
  G$, so there exist points $\zeta_-$ and $\zeta_+$ in $\boundary G$ and a
  sequence $(g_n)_{n = 1}^\infty$ of elements of $G$ such that:
  \begin{enumerate}
    \item the sequence $g_n\cdot\xi$ converges to $\zeta_-$ as $n$ tends to
      infinity, and
    \item the sequence $(g_n)$, considered as continuous maps from $\boundary
      G$ to $\boundary G$, converges uniformly on compact subsets of $\boundary
      G - \set\xi$ to the constant map with image $\zeta_+$.
  \end{enumerate}
  See~\cite{bowditch99c} for further explanation of this property.

  Let $U$ be a neighbourhood of $\zeta_+$ such that the closure of $U$ does not
  contain $\zeta_-$ and let $V$ be a neighbourhood of $\zeta_+$ contained in
  $U$ such that any map from the circle to $V$ is nullhomotopic in $U$. Then by
  the properties of the sequence $(g_n)$ described above, there exists an
  element $g_n$ such that $g_n\cdot\xi \notin U$ and $g_n\composed\gamma$ has
  image in $V$. Then $g_n\composed\gamma$ admits a contracting homotopy with
  image in $U$, and the image of this homotopy under $g_n^{-1}$ is a
  contracting homotopy of $\gamma$ with image contaed in $\boundary G -
  \set\xi$.
\end{proof}

If $G$ is a group such that $\boundary G - \set\xi$ is disconnected for some
point $\xi$ in $\boundary G$, arguments due to Bowditch~\cite{bowditch98b} and
Swarup~\cite{swarup96} build an action of $G$ on a tree-like space to show that
$\boundary G$ is disconnected. It follows that $\boundary G$ is locally path
connected if and only if either $G$ is one- or two-ended. No such argument is
available in the higher dimensional case, but on the other hand no known
example of a boundary of a hyperblic group contradicts the corresponding
assertion. Therefore the following questions are natural.

\begin{question} Is there a one-ended hyperbolic group $G$ such that $\boundary
  G$ is simply connected but not locally simply connected? Conversely, if
  $\boundary G$ is locally simply connected but not simply connected, does it
  follow that $\boundary G$ is a circle and $G$ is virtually Fuchsian?
\end{question}

It is also natural to ask whether the theorem of this paper generalises to
higher dimensions. The difficulty is in generalising the proof of
Proposion~\ref{prop:sufficient_condition}: given a sequence of maps from $S^0$
to $\boundary G$ we pass to a uniformly convergent subsequence, but this is not
possible for higher dimensional spheres without controlling the modulus of
continuity of the maps.

\begin{question} Suppose that $\boundary G - \set\xi$ is $k$-connected for any
  point $\xi \in \boundary G$. Does it follow that $\boundary G$ is
  locally-$k$-connected?
\end{question}
